\documentclass{article}

%1.  a statistician's dream environment
%2.  what emacs has to offer
%        a) buffers
%        b) key-binding
%        c) other packages
%                1) font-lock
%                2) shell/comint
%                3) ange-ftp/EFS/Tramp
%3. what ESS has to offer
%        1) font-lock/key-binding/other packages revisited
%        2) statistical process interaction
%                a) interactive
%                        1) R/S
%                        2) SAS
%                b) batch
%                        1) SAS
%                        2) BUGS
%4. ESS as an open source project
%        1) early days
%                a)S-mode
%                b)SAS-mode
%        2) transition
%                a)ESS-mode
%                b)Emacs/XEmacs
%        3) modern era
%                a)GPL, cvs, ssh
%                 b)ESS 6, the next generation?


%\renewcommand{\baselinestretch}{1.5}
\addtolength{\oddsidemargin}{-0.5in}
%\addtolength{\topmargin}{-0.2in}
\addtolength{\textheight}{0.2in}
\addtolength{\textwidth}{1in}

%%%
\usepackage[authoryear,round]{natbib}
%or (if you have an unshiny latex installation)
%\newcommand{\citep}[1]{{\{\sf#1\}}}
%%%
\usepackage{alltt}

%% Postscript fonts
%\usepackage{times}
\usepackage{graphicx}

\ifx\pdfoutput\undefined
  %% Stuff wout hyperref
  \def\url#1{\textsf{#1}} % To help fit in lines
\else
  %% Stuff with hyperref
  \usepackage{hyperref}
  %%\hypersetup{backref,colorlinks=true,pagebackref=true,hyperindex=true}
  \hypersetup{backref,colorlinks=false,pagebackref=true,hyperindex=true}
\fi

%%---End of package requiring ---------- Own Definitions -------------

\newcommand*{\SAS}{\textsc{SAS}{\textregistered} }
\newcommand*{\Splus}{\textsc{S-Plus}}
\newcommand*{\XLispStat}{\textsc{XLispStat}}
\newcommand*{\Stata}{\textsc{Stata}}
\newcommand*{\Rgui}{\textsc{Rgui}}
\newcommand*{\Perl}{\textsc{Perl}}
\newcommand*{\Fortran}{\textsc{Fortran}}
\newcommand*{\Scmt}[1]{\hbox{\qquad {\footnotesize \#\#} \textsl{#1}}}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}{Example}[section]

\newcommand{\stexttt}[1]{{\small\texttt{#1}}}
\newcommand{\ssf}[1]{{\small\sf{#1}}}
\newcommand{\elcode}[1]{\\{\stexttt{\hspace*{2em} #1}}\\}
\newenvironment{Salltt}{\small\begin{alltt}}{\end{alltt}}
\newcommand{\US}{{\char'137}}        % \tt _
\newcommand{\marpar}[1]{\marginpar{\raggedright#1}}
\newcommand{\file}[1]{`\stexttt{#1}'}

%%--------------------------------------------------------------- Start Text

\title{Emacs Speaks Statistics:  A Rapid Application Development (RAD)
  Environment for Statistical Analysis Packages}

\author{A.J. Rossini \and Martin M{\"a}chler \and Kurt Hornik \and Richard
  M. Heiberger \and Rodney Sparapani \footnote{%
%%
    A.J. Rossini is Research Assistant Professor in the Department of
    Biostatistics, University of Washington and Joint Assistant Member at
    the Fred Hutchinson Cancer Research Center, Seattle, WA, USA
    (E-mail: rossini@u.washington.edu);
%%
    Martin M{\"a}chler is Senior Scientist and Lecturer in the Seminar for
    Statistics, ETH Zurich, Zurich, Switzerland
    (E-mail: maechler@stat.math.ethz.ch);
%%
    Kurt Hornik is Professor in the Institut f{\"u}r Statistik,
    Wirtschaftsuniversit{\"a}t Wien and the Institut f{\"u}r
    Wahrscheinlichkeitstheorie und Statistik, Technische Universit{\"a}t
    Wien, Vienna, Austria (E-mail: Kurt.Hornik@r-project.org);
%%
    Richard M. Heiberger is Professor in the Department of Statistics at
    Temple University, Philadelphia, PA, USA (E-mail: rmh@temple.edu);
%%
    and Rodney Sparapani is Senior Biostatistician at the Medical College
    of Wisconsin, Milwaukee, WI, USA (E-mail: rsparapa@mcw.edu)}}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}
  Emacs Speaks Statistics (ESS) provides a user interface for
  developing statistical applications and performing data analysis
  using any of several common statistical programming languages.  ESS
  is based on the Emacs text editing environment and is a member of
  the programming tools category of Integrated Development
  Environments (IDEs), which are tools which aid in developing and
  visualizing computer applications.  We discuss how it works, why one
  would consider using it, and additional extensions for increasing
  statistical programming efficiency.
\end{abstract}

\noindent Keywords: Data Analysis, Programming Tools, User Interfaces, S, \SAS,
\Splus, R, \XLispStat, \Stata, BUGS

\baselineskip=2pc

\section{Introduction}
\label{sec:introduction}

Common tasks faced by a statistician are not much different from those
faced by software developers, since a great deal of statistical
activity involves some form of computer programming.  Many
statistical research activities, such as data analysis and
communication, are enhanced by computers.  This puts the user
interface, which acts as a buffer between the statistician and
computer, in the critical role of facilitating or hindering these
tasks.  This paper introduces Emacs Speaks Statistics (ESS),
which is a set of extensions to the Emacs text editor which facilitate
the use of a number of common statistical programs and languages by
providing shortcuts and extensions for often-performed tasks.

%% Does it make sense to add an additional paragraph on "history of
%% statistical interfaces from programming it yourself to command line
%% (SPSS/SAS/BDMP in the 80s), to spreadsheet-style interfaces
%% (spreadsheets, AND minitab, spss now, etc), etc?  This would be the
%% 2nd paragraph, if so. 
%% 
%% I don't think it is necessary, although, I'm not entirely opposed.
%% AFAIC, anyone who wants to use ESS will not need to know all of
%% that per se.  They will just need to know what works today IMO.
%% Rodney

ESS provides an interface to statistical packages, and provides,
through Emacs, additional tools which facilitate both statistical
software development and data analysis.  Currently supported
statistical packages are the following:  the S family of languages (which include
S \citep{BecRCW88,ChaJH92,ChaJ98}, \Splus{} \citep{Splus}, and R
\citep{ihak:gent:1996}); \SAS \citep{SAS:8}; \XLispStat\ 
\citep{Tier90} and its extensions Arc \citep{Cook:Weisberg:1999} and
ViSta \citep{youn:fald:mcfa:1992}; \Stata\ \citep{Stata:6.0}; Omegahat
\citep{DTLang:2000}; and BUGS \citep{BUGS}.  ESS can be
extended to accommodate most statistical packages which provide either
an interactive command-line or batch file processing.

The present paper provides historical and non-technical information on
Emacs and ESS.  A general introduction and usage instructions can be
found in \cite{heiberger:dsc:2001}.  The documentation that comes with
ESS provides details of its implementation and examples of its use.
We first describe the Emacs text editor, the platform on which ESS is
built.  That is followed by the details of ESS and how it facilitates
common statistical computing activities.  We then describe the unique
history that ESS has had as an open-source, free-software project.  
It has been passed down by generations of developers and has 
united various Emacs packages which support
statistical languages and programming.  Finally, we conclude with
future tasks and extensions.

\section{Emacs}
\label{sec:emacs}

Emacs is a mature, powerful, and extensible text editing system which
is freely available under the GNU General Public License for a large
number of platforms, including most flavors of Unix, Microsoft Windows
and Apple Macintosh.  There are two open-source, free-software
implementations of Emacs (GNU Emacs \citep{GNU Emacs} and XEmacs
\citep{XEmacs}).  Emacs shares many features with
word processors, and some characteristics with operating systems.
Most importantly, Emacs can interact with and control other programs.

Emacs was originally written when character-based terminals with only
a keyboard for input was the most advanced method of access to
computers.  Common Emacs commands were assigned to key sequences,
which is known as key-mapping or key-binding.  These days, Emacs is
rarely used within a terminal or terminal-emulator, but key-mapping of
commands is still an important time-saving feature.  Over the last
decade, Emacs has been extended to use graphics-based windowing
systems such as X11, Microsoft Windows, and Apple Macintosh.  These
windowing systems provide a graphical user interface (GUI) which
allows the user to interact with either a keyboard or a mouse and for
multiple applications to share a single display.  Emacs menus and
toolbars allow mouse access to frequently used actions; for instance,
when you don't know or can't remember a key-mapping.  User-defined
menus and toolbars can be constructed as needed.

Emacs provides facilities which go beyond ordinary text editing.  
When opening a file, Emacs can complete partially typed 
file names by referring to the typed directory.  This file is placed
in a space created within the Emacs application called a buffer.
Buffers can be used for more than just editing text files; for instance,
they can be used for controlling other programs or viewing the contents of
a directory.  Buffers allow the user to simultaneously edit multiple files
and concurrently control multiple programs simply by switching between buffers.

Emacs capabilities are extended by text files containing Emacs Lisp,
(shortened as elisp), a dialect of Lisp
\citep{RChassell1999,PGraham:1996}.  Emacs commands, written in elisp,
can be called interactively by name or by their associated key-map.
Emacs functions are also written in elisp, but they can only be called
by Emacs commands or other Emacs functions rather than directly by
name or key-maps.  The most important extensions to Emacs are called
modes.

Major-modes provide two kinds of support.
First, a customized environment (key-maps, capabilities) for file
types which are determined from the file name's extension, i.e. the characters
at the end of the file name that follow a period like ``txt'', ``s''
or ``sas''.  When dictated by the file name, you can only utilize one 
major-mode at a time.  An example of this kind of major-mode is ESS 
(technically, the name is actually \stexttt{ess-mode}).  Major-modes of this
kind provide automatic indentation and navigation in units of characters,
words, lines, sentences, paragraphs, and pages as well as many other
features.  The second kind of major-mode is for running other programs.  
In this case, the programs standard input can be drawn from a buffer
or a file and its standard output can be directed to a buffer or a file.  
These buffers can control programs that transfer files over a network,
perform version control operations or many other tasks.
An important example of this kind of major-mode is \stexttt{shell-mode}.

Minor-modes, by contrast, provide
services that are not mutually exclusive.  This includes emulation of
another editor or cooperating with a version control program.  For
example, the minor-mode \stexttt{font-lock-mode} allows Emacs to highlight, with
fonts or colors, the syntax of a programming language whose
characteristics are described by a major-mode like ESS. 
Remote editing of files is provided by minor-modes \stexttt{ange-ftp} and 
\stexttt{EFS} for ftp access or \stexttt{tramp} for scp/ssh/telnet.
Interacting with version control programs is provided by the minor-mode
\stexttt{vc-mode} for packages like CVS, RCS and SCCS. 

Most programming and documentation tasks fall under the realm of text
editing.  This work can be enhanced by modes which provide features such 
as contextual highlighting and recognition of special reserved words 
appropriate to the programming language in use.  Emacs supports
folding, outlining, and bookmarks which can assist with maneuvering
around a file.  Emacs shares many features with word-processing
programs and cooperates with document preparation systems
that perform type-setting such as \LaTeX or hypertext markup
like HTML and XML.
%XSLT, and Noweb.  Type-setting and word-processing, which
%focus on the presentation of a document, are tasks that are not pure
%text-editing.

%insertion and deletion: viewing two or more files at once; editing formatted text;
%visual comparison of two similar files; and navigation 
%in units of characters, words, lines, sentences, paragraphs, and pages.  

 Emacs handles the
interface to both source code and transcripts contextually, providing
syntax highlighting, bookmarking features, interfaces to directory
structure, and command-history.  Other extensions to Emacs allow it to
act as a World-Wide-Web browser, a highly sophisticated mail and news
reader, a shell/terminal window with history, and as an interface to
other common text-based tools such as spell checking programs.  The
Emacs keyboard and mouse interface can be re-mapped to resemble that
of other text-editors, such as vi, wordstar, and brief.


\section{ESS extends Emacs}
\label{sec:ess-extends-emacs}

Emacs' power and flexibility make it a sensible starting choice to
build an interface for statistical analysis packages.  ESS extends
Emacs to provide a functional, easily extensible and uniform interface
for multiple statistical packages.  This is done in two ways.  First,
shortcuts and features for accelerated editing of files have been
implemented.  Second, interaction with the statistical
language is split by function; general, obtaining help, programming.
\textbf{THIS NEEDS WORK}.

%This is done by extending the editor to provide
%additional useful features, and in the case of interactive statistical
%programs, sitting ``in front of'' their CLI and intercepting and
%modifying I/O as needed.  We describe the exact features more in this
%section.  

%% change/edit
%Figure~\ref{fig:1} provides an example of how it looks when being used
%with XEmacs.  This screen-shot shows both SAS and R code being examined
%at the same time, with an R interactive process being controlled from
%within XEmacs.

%Figure~\ref{fig:2} shows ESS running in NTemacs 21.0 on Microsoft
%Windows 2000.  We are currently displaying 6 buffers.
%\stexttt{highlight.s} illustrates several uses of syntactic
%highlighting.  The most glaring one is the bright purple indicator for
%the unbalanced parentheses.  We also see color choices for keywords
%(\stexttt{if}), comments, and quoted strings.  The string on the last
%line was not properly terminated and we are immediately warned by the
%string color staying on through (what we think is) the end of the
%line.

%The two buffers \stexttt{transcript-before.st} and
%\stexttt{transcript-after.st} show transcript editing.  A single ESS
%command converted the before buffer into the after buffer by removing
%all lines that do not begin with a prompt character.

%The last three buffers show that a single S language source file can
%be used with two (or more) executing processes.  In this example we
%sent over first a subset of the line in \stexttt{tmps.s} and then the
%entire line to the instance of S+4 running in an inferior
%\stexttt{iESS(Sqpe)} buffer.  Then we switched the connection to send
%the same line to the instance of R running in the \stexttt{iESS(R)}
%buffer.  The dialog about the process switch appears in the message
%line.


\subsection{Features and capabilities}
\label{sec:ESS:features}


ESS strongly supports the S family (S, \Splus, and R).  
\SAS is also well supported.  \Stata\ 
and \XLispStat\ (and the \XLispStat\ extensions, ARC and ViSta) are
supported with the basic functionality of syntax highlighting and
process-interfacing.  ESS and its interface are fully discussed in the
next section.

%% Rodney:  I've tried to rearrange these in a more coherent fashion
%% but, I'm hindered by the fact that I don't use some of these
%% features.  Can someone help?

\paragraph{Syntactic indentation and color/font-based source code
  highlighting.}  The ESS interface includes a description of the
syntax and grammar of each statistical language it knows about.
%%  This gives ESS the ability to edit the programming language code, 
%% often more smoothly than with editors distributed with the languages. 
The process of programming code is enhanced as ESS provides the user with
a clear presentation of the code with syntax highlighting to denote
assignment, reserved words, presence of strings, and comments.  ESS
has customizable automatic indentation, with the customization based
on the syntactic structure of groups of expressions.  ESS knows the
structure of transcripts of the executing session and provides
syntactic highlighting for transcripts.  

%% Rodney:  What is source code checking?  Should we lump this with highlighting?
\paragraph{Source code checking.}
ESS facilitates the editing of source code by providing a means for
loading and error-checking of small sections of code for S,
\XLispStat, and \SAS.  This allows for source-level debugging of batch
files.

%% Rodney:  What does ``entire function definitions'' mean?
\paragraph{Partial code evaluation.}
Emacs can send individual lines, entire function definitions, marked
regions, and whole edited buffers from the window in which the code is
displayed for editing to the statistical package for
execution.  Emacs sends the code directly to
the statistical package and receives the output, if any, which is 
the result.  This is a major improvement over cut-and-paste as it does
not require switching between buffers or windows.  The response is received
immediately in an editable Emacs buffer.

\paragraph{Interactive transcripts and history of commands.} 
ESS interacts well with the
executing statistics language/program and provides means for searching
the command-line history for previous commands and editing them for
current use.

%% Rodney:  should we merge these transcript sections?
\paragraph{Transcription Editing and Reuse.}
Once a transcript log is generated, perhaps by saving an \file{iESS}
buffer, transcript-mode assists with reuse of part or all of the
entered commands.  It permits editing and re-evaluating the commands
directly from the saved transcript.  This is useful for
demonstration of techniques as well as for reconstruction of data
analyses.  There currently exist functions within ESS for cleaning
transcripts from S languages back to source code by finding all input
lines and isolating them into an input file.

%% Rodney:  inferior is about as opaque as it gets.  Let's say
%% ``interactive'' instead.  And, I created another interactive
%% section above.  Should we lump these.
\paragraph{Process interaction.}
Emacs has historically referred to processes under its control as
``inferior'', accounting for the name inferior ESS (\stexttt{iESS}) to
denote the mode for interfacing with the statistical package.  The
output of the package goes directly to an editable text buffer in Emacs.
This mode allows for command-line editing and saving history, as well as
recalling and searching for previously entered commands.  Filename
completion is available.  In addition (currently only for S languages),
there exists object-name and function-name completion.  Transcripts are
easily recorded and can be edited into an ideal activity log which can
then be saved.  There is a good interface for handling and intercepting
calls to the internal help systems for S, \XLispStat, and \Stata.

%% Rodney:  actually, I've not been able to get this to work with
%% ange-ftp, EFS or Tramp.  The situation may have been different
%% 3 or 4 years ago, but with the security crackdowns due to 
%% flaws in telnet and ftp; this is not likely to work for most
%% people.  I suggest that we either drop this section or move
%% it to future work so we can nail it down.
\paragraph{Interacting with statistical programs on remote computers.}
ESS provides the facility to edit and run programs on remote machines
in the same session and with the same simplicity as if they were
running on the local machine.  The remote machine could be a very
different platform than the local machine.

\paragraph{Object name completion.}
For languages in the S family,
ESS provides object-name completion of both user- and
system-defined functions and data.  ESS can dump and save objects
(user- and system-generated) into text files in a formatted manner for
editing, and reload them (possibly after editing).
%% back into the statistical language/program.

\paragraph{Help File Editing (R).}
ESS provides an interface for writing help files for R functions
and packages.  It provides the ability to view and execute embedded R
source code directly from the help file in the same manner as ESS
normally handles code from a source file.  \stexttt{Rd} mode provides
syntax highlighting and the ability to submit code to a running ESS
process, either R or \Splus.


\subsection{ESS Facilitates Common Tasks}
\label{sec:ess-facil-comm}

\subsubsection{Multiple Tools}
\label{sec:multiple-tools}

Basically, statistical packages are either intended for general 
statistical analyses or more specialized forms of statistical analyses.  
The specialized statistical packages can be far more efficient for
specialized statistical analyses, but this is balanced by their inability
to perform a wide range of general statistical analyses.  Tightly coupled 
inter-operability between general and specialized packages rarely exists, 
but such a facility is often desired.  For example, a general purpose 
package such as R does not perform Bayesian analyses as easily as BUGS does.  
On the other hand, BUGS lacks breadth in the range of analyses and results 
it can generate.  For this reason, BUGS is often distributed with R packages, 
like CODA and BOA, for importing and analyzing the results in R.  Another
point of contention is the difference in the interfaces between general
packages and specialized packages.  For example, the interfaces for BUGS 
and R are somewhat similar, but there are enough differences to create confusion.

ESS is an extension package for
the Emacs editor which provides a single interface for a variety of
statistical computing tasks.  ESS is optimized for statistical programming
and interactive data analysis.  Statistical programming is the writing of
computer programs for data analysis and processing.  These programs might be
written in a computer language that requires as a compiler such as C or 
\Fortran.  But, more likely, they are written in a  computer
language that only requires an interpreter such as R, \SAS or \XLispStat.
%%An interpreted language is executed by an interpreter, but a compiled language
%%must be processed by a compiler which creates an executable form of the
%%program that can be executed.  There is a trade-off in time between interpreted and compiled
%%programs.  For interpreted languages you save time by not having to compile
%%the program, but the program will not run as fast as if it had been compiled.
%%Another advantage of intepreted languages is the support they provide for
%%specialized tasks.  For example, C and \Fortran don't provide you with
%%the statistical analysis capabilities that R, \SAS or \XLispStat
%%do.  Compiled programs and interpreted programs can be used in conjunction.
%%For instance, a compiled C function could be called from an interpreted 
%%language like R to speed up the calculations.
Entering commands for interactive data analysis is a similar activity.
In either case, text is written in a computer language and sent to a
computer program for evaluation. 
%%The primary difference is that the
%%results of a small set of commands are of critical interest for review
%%in the analysis phase, but the results of all commands are of interest
%%in the coding phase.  

\subsubsection{Conflicts in keymappings}
\label{sec:confl-keym}

Simple conflicts between interfaces are exemplified by different
keystrokes for editing tasks such as cut, copy, paste, beginning of
line, end of line, etc.  These may be the most aggravating
because our fingers are acting without our brains having to
specifically think about it, but differences in interfaces
circumvent this learned behavior.  ESS solves this problem by
providing a uniform keyboard interface.

Complex conflicts between interfaces can involve the coordination of
several data files, multiple statistical software packages, and the
corresponding source code in each of these languages, combined for a
single analysis.  ESS, as part of Emacs, has tools which assist in
this, including support for version and source code control systems,
tools for accessing programs or files on remote machines, and
interfaces to documentation systems including \LaTeX\ and
\textsc{xml}.  In addition, Emacs can assist with, or be programmed to
perform, many tasks related to data cleaning, management, and editing.


\subsubsection{Concurrent Use of Multiple Machines and Operating
  Systems}
\label{sec:conc-use-mult}

It can be useful to have multiple statistical processes running
simultaneously, either on a single machine or a variety of machines.
This capability assists with code design and testing across multiple
versions of statistical software packages, as well as large scale
numerical simulations.  For example, one might want to be connected to
multiple R processes of different versions in order to verify behavior
on different versions of the same software; multiple processes of the
same version to perform test-and-run scenarios where one process is
doing long-term processing while the other is doing short-term
testing; simulations; and to distribute the load over a variety of
remote machines.

\section{History of ESS}
\label{sec:ESS:history}

ESS is an example of how open-source products can continue to
develop beyond what their initial authors planned.  To date, there have been 3
generations of developers and hopefully, there will be many more to come. 
%%and the genealogy is more tree-like thanlinear.  
There were 2 modes from which ESS descended:  S-mode and SAS-mode.

%% ESS originated as S-mode, which provided an interface for
%% programming and process control under GNU Emacs for \Splus\ version 3,
ESS was brought to life in 1989 initially as S-mode, to
provide extensions to GNU Emacs for editing S and \Splus{} files.  This
was written by Doug Bates, Ed Kademan, Frank Ritter, and
Mike Meyer.  David Smith was the next primary maintainer, and his most
important contribution was to enhance the interface to the \Splus{}
command-line interface and allow for control of multiple processes
simultaneously.  In 1995, A.J.  Rossini extended S-mode to support
XEmacs.  At the
same time, extensions from ETH, written by Martin M{\"a}chler, were folded
into what was to become the primary codebase.  By 1996,
S-mode was available for both GNU Emacs and XEmacs and supported S, \Splus, and
to some extent R.  Kurt Hornik was instrumental in enhancing support
for R, in particular, providing Rd mode.

The other ancestor of ESS is SAS-mode,
written by Tom Cook, which began in 1994 and was partially based on
GNU Emacs macros written by John Sall \textbf{NEED TO GET
  DATES FROM AJR's ARCHIVE!?}.  Rossini extended SAS-mode to work with
XEmacs in 1995, and this provided the initial
impetus for a uniform codebase for statistical programming.
The extension to a language-independent, generic interface was further prompted
by the success of R and the desire for more complete R support.  This led to a merger
of SAS-mode and S-mode along with changes 
to accommodate multiple languages in a flexible way.

During 1996 and 1997, Richard M. Heiberger further incorporated
SAS-mode into ESS and designed the inferior ESS mode for \SAS.  In
1997, more changes were made to facilitate  
%% the grand redesign of the internals occurred, where by a generic
%% means for 
configuring ESS for new statistical languages like \Stata and \XLispStat.
Most of the work, until 1998, was done for Unix-based statistics packages.
The first example of interprocess communication for Windows using DDE
was constructed in 1998 by Brian Ripley.  Based on this work, Heiberger provided 
interfaces for Windows versions of \Splus.

In 1998, Rodney Sparapani designed \SAS batch file processing for ESS.
Sparapani and Heiberger developed \SAS batch support including function
key behavior that follows \SAS{\'s} own definitions,
a more comprehensive and efficient syntax highlighting scheme,
automated error message lookup in the \SAS log file and extended the
\stexttt{*ESS-elsewhere*} functionality to include \SAS.  This provides
a powerful development environment for \SAS.
%% Rodney:  I'm still not comfortable with placing SAS-elsewhere as 
%% a done deal.  I think this should be a future direction.
%% We could say something like:  we are currently working on remote
%% file access with respect to ESS and there should be a new
%% release by the time you read this.  Comments?

In 2001, Sparapani added BUGS batch file processing for ESS.  This 
includes support for syntax highlighting and template-based source file
generation that provides the capability of specifying all the necessary
parameters for a BUGS batch run in a single file.

\section{Discussion}
\label{sec:discussion}

There are two active areas of extensions for user environments.  One
is to enhance the capabilities of the IDE for statistical practice;
this includes implementing such common IDE features as object
browsers, tool-tips, and interfacing cleaning.  The other is to target
appropriate potentially useful programming methodologies for transfer
to statistical practice.

Literate Programming methodologies \citep{Knuth:1992,NRamsey:1994} are
a natural fit for statistical practice.  We refer to the application
to statistical analysis as Literate Statistical Practice
\citep{rossini:dsc:2001}.  The tools used are Noweb
\citep{NRamsey:1994} and either \LaTeX, \textsc{html}, or \textsc{xml}
for documenting and explaining the analysis.  This approach to
programming encourages the use of a literary documentation style to
explain the programming code for the data analysis.  The program can
then be extracted from the documentation text for realizing the
statistical analysis.

Important IDE extensions which should be implemented in future
versions include class browsers, analysis templates, tool-tips, and
similar features.  Class browsers can be thought of as a tree or
outline for presenting datasets, variables and functions in the
context of what they represent; this allows for rapid and appropriate
inspection.  Analysis templates would allow statistics centers and
groups to provide standardized templates for initiating an analysis.
While most IDE features have been developed for object-oriented
languages, the above also can apply to non-object oriented
programming.

ESS is one of the first IDEs intended for statisticians.  It provides
an enhanced, powerful interface for efficient interactive data
analysis and statistical programming.  It is completely customizable
to satisfy individual desires for interface styles as well as being
extensible to additional statistical languages and analysis packages.


\bibliographystyle{plainnat}
%\pdfbookmark[1]{References}{section.7}
\bibliography{essJCGSv2}


\end{document}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
